---
title: "Lab 2: Logistic Regression"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE, 
                      include = TRUE,
                      quiet = TRUE)
```

# Learning Objectives - Explore 
Exploratory Data Analysis (cont’d):
- Pairs plot to show correlation between variables and avoid multicollinearity (see 8.2 Many predictors in a model)
Logistic Regression seen as an evolution of techniques
- Linear Model to show simplest multivariate regression, but predictions can be outside the binary values.
- Generalized Linear Model uses a logit transformation to constrain the outputs to being within two values.
- Generalized Additive Model allows for “wiggle” in predictor terms.
- Maxent (Maximum Entropy) is a presence-only modeling technique that allows for a more complex set of shapes between predictor and response.

# Load packages and environmental data layers  
```{r}
librarian::shelf(
  DT, dplyr, dismo, GGally, here, readr, tidyr)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = F)

dir_data    <- here("data")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

pts_env <- read_csv(pts_env_csv)
nrow(pts_env)
```

## Displate environmental data in a table 
```{r}
datatable(pts_env, rownames = F)
```
## Use `GGally` to look at pair plots and examine correlations between environmental variables 
```{r}
GGally::ggpairs(
  select(pts_env, -ID),
  aes(color = factor(present)))
```
# Logistic Regression 
## Setup Data 
- drop rows of data with any NA values (later, we'll learn how to impute values)
- remove terms we don't want to model 
```{r}
d <- pts_env %>% 
  select(-ID) %>%  # remove terms we don't want to model
  tidyr::drop_na() # drop the rows with NA values
nrow(d)
```

## Linear Model 
- dependent variable Y: presence/absence, aka 1/0
- independent variables X: everything else in the dataframe, aka the environmental layers 
```{r}
mod <- lm(present ~ ., data = d)
summary(mod)
```

Note: a linear model is ineffective because it predicts values outside the 0 - 1 range
```{r}
y_predict <- predict(mod, pts_env, type="response")
y_true    <- pts_env$present

range(y_predict)
range(y_true)
```

# Generalized Linear Model 
To solve this problem, we will apply a Logit Transformation 
```{r}
# fit a generalized linear model with a binomial logit link function
mod <- glm(present ~ ., family = binomial(link="logit"), data = d)
summary(mod)
```
Note: we are not within the 0 - 1 range we want to be in
```{r}
y_predict <- predict(mod, d, type="response")
range(y_predict)
```

## Term Plots 
```{r}
termplot(mod, partial.resid = TRUE, se = TRUE, main = F)
```

# Generalized Additive Model (GAM)
We can further improve the GLM by adding "wiggle" to the relationship between the predictor and response variables 
```{r}
librarian::shelf(mgcv)

# fit a generalized additive model with smooth predictors
mod <- mgcv::gam(
  formula = present ~ s(WC_alt) + 
                      s(WC_bio1) + 
                      s(WC_bio4) + 
                      s(WC_bio12) + 
                      s(ER_climaticMoistureIndex) + 
                      s(ER_tri) + 
                      s(ER_topoWet) + 
                      s(lon) + 
                      s(lat), 
  family = binomial, data = d)
summary(mod)
```

## Term Plots 
```{r}
plot(mod, scale=0)
```

# Maxent (Max Entropy)
This is the most commonly used species distribution model because it requires few input data points, all of which can be presence observation points, and is easy to use with a Java GUI. 
```{r}
# load extra packages
librarian::shelf(
  maptools, sf)

# show version of maxent
maxent()
```

```{r}
# get environmental rasters
# NOTE: the first part of Lab 1. SDM - Explore got updated to write this clipped environmental raster stack
env_stack_grd <- file.path(dir_data, "env_stack.grd")
env_stack <- stack(env_stack_grd)
plot(env_stack, nc=2)
```

```{r}
# get presence-only observation points (maxent extracts raster values for you)
obs_geo <- file.path(dir_data, "obs.geojson")
obs_sp <- read_sf(obs_geo) %>% 
  sf::as_Spatial() # maxent prefers sp::SpatialPoints over newer sf::sf class

# fit a maximum entropy model
mod <- maxent(env_stack, obs_sp)
```

```{r}
# plot variable contributions per predictor
plot(mod)
```

# Term Plots 
```{r}
# plot term plots
response(mod)
```
## Predict 
```{r}
# predict
y_predict <- predict(env_stack, mod) #, ext=ext, progress='')

plot(y_predict, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

