---
title: "Lab 1d: Species Distribution Modeling - Evaluate Models"
author: "Mia Forsline"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE, 
                      include = TRUE,
                      quiet = TRUE)
```

# Learning Objectives 
- Evaluate model performance
- Calibrate model parameters 

# Setup 
- load packages
  - `dismo` - species distribution modeling: `maxent()`, `predict()`, `evaluate()`
  - `usdm` - uncertainty analysis for species distribution models: `vifcor()`
- set options
- set `ggplot` theme 
- set directory paths 
- read in the data
  - observation points
  - raster stack of environmental layers 
```{r}
librarian::shelf(
  dismo, # species distribution modeling: maxent(), predict(), evaluate(), 
  dplyr, ggplot2, GGally, here, maptools, readr, 
  raster, readr, rsample, sf,
  usdm)  # uncertainty analysis for species distribution models: vifcor()
select = dplyr::select

set.seed(42)
options(
  scipen = 999,
  readr.show_col_types = F)
ggplot2::theme_set(ggplot2::theme_light())

dir_data      <- here("data")
pts_geo       <- file.path(dir_data, "pts.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")
mdl_maxv_rds  <- file.path(dir_data, "mdl_maxent_vif.rds")

pts <- read_sf(pts_geo)
env_stack <- raster::stack(env_stack_grd)
```

## Split observation data into training and testing sets
- the training data set uses 80% of the data 
- the testing data set uses the remaining 20% of the data 
```{r}
pts_split  <- rsample::initial_split(
  pts, prop = 0.8, strata = "present")
pts_train  <- rsample::training(pts_split)
pts_test   <- rsample::testing(pts_split)

pts_train_p <- pts_train %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_train_a <- pts_train %>% 
  filter(present == 0) %>% 
  as_Spatial()
```

# Calibrate: Model Selection 
- use a pairs plot to check correlation between environmental variables and assess multicollinearity issues 
```{r}
pairs(env_stack)
```

## Calculate variance inflation factor per predictor, a metric of multicollinearity between variables
```{r}
vif(env_stack)
```

## Stepwise reduce predictors, based on a max correlation of 0.7
- conduct a multicollinearity reduction using `vifcor()`
```{r}
v <- vifcor(env_stack, th=0.7) 
v
```

## Reduce the environmental raster stack by excluding variables that were too correlated 
- visualize the remaining variables with another pairs plot 
```{r}
env_stack_v <- usdm::exclude(env_stack, v)

pairs(env_stack_v)
```

## Fit a maximum entropy model using `maxent()`
- plot variable contributions per predictor
```{r}
pts_train_test <- pts_train %>% 
  select(present, geometry)

if (!file.exists(mdl_maxv_rds)){
  mdl_maxv <- maxent(env_stack_v, pts_train)
  readr::write_rds(mdl_maxv, mdl_maxv_rds)
}
mdl_maxv <- read_rds(mdl_maxv_rds)

plot(mdl_maxv)
```

## Plot term plots
```{r}
response(mdl_maxv)
```

## Predict 
```{r}
y_maxv <- predict(env_stack, mdl_maxv) #, ext=ext, progress='')

plot(y_maxv, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

# Evaluate 
## Area Under the Curve (AUC), Receiver Operator Characteristic (ROC) Curve, and Confusion Matrix 
- split the test data into presence and absence
- use `dismo()` to calculate the AUC and max TPR + TNR
```{r}
pts_test_p <- pts_test %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_test_a <- pts_test %>% 
  filter(present == 0) %>% 
  as_Spatial()

y_maxv <- predict(mdl_maxv, env_stack)
#plot(y_maxv)

e <- dismo::evaluate(
  p     = pts_test_p,
  a     = pts_test_a, 
  model = mdl_maxv,
  x     = env_stack)
e
```
## Plot the ROC Curve 
- identify the threshold and mark it in blue on the plot 
```{r}
plot(e, 'ROC')

thr <- threshold(e)[['spec_sens']]
thr
```

## Create the Confusion Matrix 
```{r}
p_true <- na.omit(extract(y_maxv, pts_test_p) >= thr)
a_true <- na.omit(extract(y_maxv, pts_test_a) < thr)

# (t)rue/(f)alse (p)ositive/(n)egative rates
tpr <- sum(p_true)/length(p_true)
fnr <- sum(!p_true)/length(p_true)
fpr <- sum(!a_true)/length(a_true)
tnr <- sum(a_true)/length(a_true)

matrix(
  c(tpr, fnr,
    fpr, tnr), 
  nrow=2, dimnames = list(
    c("present_obs", "absent_obs"),
    c("present_pred", "absent_pred")))
```
```{r}
plot(e, 'ROC')

thr <- threshold(e)[['spec_sens']]
thr

points(fpr, tpr, pch=23, bg="blue")
```


## Plot the areas where we predict *Necturus maculosus* presence 
```{r}
plot(y_maxv > thr)
```

